Safe Transferable Regions

### 2015-11-18 paper/type-inference.tex
Edits to Sec.4. Formatting changes.

We formalize region type inference as an elaboration function from
programs in $\absof{\FB}$ (i.e., region-erased \FB) to programs in
\FB.  The elaboration proceeds in four steps. In the first step we
make use of the observation that region types are \emph{refinements}
of FGJ types with region annotations and constraints over such region
annotations, and compute polymorphic region type templates for methods
and classes from their FGJ types. The templates contain region
variables ($\rho$) to denote unknown region annotations, and predicate
variables ($\varphi$) to denote unknown constraints over such region
annotations. Free region variables are generalized in types (hence,
polymorphic). Second, we make use of the computed region type
templates to elaborate expressions by introducing region variables to
denote unknown region arguments in \C{new} expressions, method calls
and function applications. While elaborating expressions, we also
build a system of constraints that capture well-formedness
requirements and subtyping relationships between type templates that
must hold (as per the static semantics in Fig.~\ref{fb-staticsem}) for
the elaboration to be valid. Third, we lift expression elaboration and
constraint generation to methods and classes. Finally, we solve the
constraints by making use of our fixpoint constraint solving algorithm
\csolvestar, which reduces the constraint solving problem to an
abduction problem in a Herbrand constraint system, and then relies on
\csolve, our abduction solver for that domain. 
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. FGJ type $\RgnZ\inang{T}$ is special; it is
always templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
region-annotated definition of \C{Pair} class from
\S\ref{sec:fb-syntax} a region type template for a method with FGJ
type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
accept one region parameter - the allocation region of their
objects.}\footnote{We abuse arrow notation to also represent types of
methods, but unlike function types, there is no allocation region
annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
region variables, and $\varphi_0$ is a fresh predicate variable
denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
$\rho$ is fresh. For a class, a template is computed in two steps. In
the first step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

The type template for a recursively defined class is computed in two
steps. First, all recursive occurances of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurances are
templatized with respect to the class template computed in the first
step, such that their region annotations are \emph{exactly} same as
the region parameters of the class. Applying this technique on a
recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
example, results in a template that uses same region variable to
denote allocation regions of the current node, and the node that the
current node refers to. 
% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
% data stored in the linked list node.  To templatize the
% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
% occurance in the type of $\C{next}$ field, and templatize the type
% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
% template of $\C{data}$ field, we compute the class's template as
% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
% \end{center} Next, we trivially templatize the type of $\C{next}$
% field in the body of the class as
% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
% resulting class represents a linked list with spine in the region
% $\rhoalloc_0$ and data objects in the region $\rho_0$.
% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
This technique can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

@@

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types in variable declarations and \C{new}
expressions with fresh region type templates, and (b). explicitly
instantiating region parameters of methods with fresh region variables
in method calls and function applications. This elaboration is
performed with respect to the polymorphic type templates of classes
and methods computed as per \S\ref{sec:fb-templatization}.

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$
%, whose corresponding $\FB$ expressions have been ascribed static
%semantics in Fig.~\ref{fig:fb-staticsem}. 
$\elabExpr$ is defined under the same context as the expression typing
judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
and $\env$ retaining their meaning. The function traverses expressions
in a syntax-directed manner of a type checker, introducing fresh
region type templates for unknown region types, while generating
constraints over region and predicate variables. The precise nature of
generated constraints is explained in \S\ref{sec:fb-constraintsem},
but in summary, they capture the relationships between the type
templates of various subexpressions and their well-formedness. Note
that $\elabExpr$ returns the region type template of the
subexpression, which is used to generate constraints for the
expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
shown) used by $\elabExpr$ implement type well-formedness and subtype
judgments from Fig.~\ref{fig:fb-staticsem}, respectively.

@@

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints. However, we do not
require region type inference to be a whole-program analysis.
Compositionality of type inference is reclaimed by solving the
constraints in a compositional fashion. Concretely, this means that
our constraint solving algorithm visits and solves every constraint
(or, every set of mutually dependent constraints) only once. It
composes computed solutions to solve other constraints that depend on
the solved constraints. Importantly, the failure to solve a dependent
constraint does not result in backtracking. We now describe our
constraint solving algorithm by demonstrating its peformance over the
example constraints shown in the previous section. A formal account of
the algorithm can be found in our technical report~\cite{techrep}.
% To simplify the presentation of our algorithm, we assume that are no
% mutually recursive definitions in the source program. Recursive
% definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occuring on the left side (resp. right side) of the turnstile is said
to occur negatively (resp. positively). A validity constraint
\emph{constrains} the set of predicate variables that occur negatively
in the constraint, while it \emph{uses} the set of predicate variables
that occur positively. A constraint is said to be \emph{recursive} if
it constrains and uses a predicate variable.
% A pair of validity constraints are said to be \emph{mutually
% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
% predicate variables that occur with opposite polarities in both the
% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
% predicate variables are also said to be mutually dependent. We call
% the transitive closure of mutual dependency relation as
% \emph{transitive dependency} relation. Like mutual dependency,
% transitive dependency is also extended to predicate variables.

@@

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) varibles in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{\cite{techrep} formally justifies the soundness of this
step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}
The constraint $c_{12}$ is a non-recursive constraint of the form
$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
this case). Such constraints are solved for $\varphi$ by computing a
maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
well-formedness condition on $\varphi$, then it is clearly a solution
for $\varphi$, because there cannot exist a weaker solution for
$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
maximally weak $\phisol$ fails the well-formedness condition, then we
establish~\cite{techrep} that it is impossible to satisfy the
constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
and check if it satisfies the well-formedness condition on $\varphi$.
The former problem can be cast as an instance of graph augmentation
problem, as described below.
\vspace*{-0.05in}
\paragraph{Definitions} Given a constraint formula $\phi$, we define
its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
vertices ($V(\phi)$) are free region variables and identifiers in
$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
in\footnote{We alternate between viewing a constraint formula ($\phi$)
as a set of constraints and a conjunction of constraints in this
description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
conjunction of symmetric outlives constraints for the purpose of graph
encoding. Conversely, given a digraph $G$, we define its constraint
encoding $\Phi(G)$ in a straightforward manner. We say that a graph
$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
exist a path between same pair of vertices in $G_1$.

@@

We have so far ignored function types in the context of type
inference. The region type system for $\FB$ admits function types with
higher-rank region polymorphism, allowing region variables to be
quantified in the types of higher-order arguments. The type system
clearly does not have the principal typing property, as demonstrated
by the \C{app2} method below:
\begin{codejava}
bool app2(A x, A y, (A $\rightarrow$ bool) f) 
    { return (f x && f y); }
\end{codejava}
There are two possibilities for the region type of \C{app2}: one,
where it allows \C{x} and \C{y} to be allocated in different regions,
but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
arguments from any region), and second, where it requires \C{x} and
\C{y} to be allocated in the same region, and allows \C{f} to be
monomorphic. None of the two types are more general than the other,
and it is not clear which one is more preferable. The concrete
manifestation of this dilemma is a constraint of form 
$\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
no additional constriants to help determine $\varphi_1$ or $\varphi_2$
before solving this constraint. Since it is always possible to
strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
does not exist unique solution to such constraints. Nonetheless, it is
possible to bias the constraint-solving algorithm to consistently
prefer one solution over the other.

>>>>

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., region annotation-erased \FB) to programs in \FB. The key
aspects of this elaboration are described below.

% We formalize region type inference as an elaboration function from
% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
% programs in \FB.  The elaboration proceeds in four steps. In the first
% step we make use of the observation that region types are
% \emph{refinements} of FGJ types with region annotations and
% constraints over such region annotations, and compute polymorphic
% region type templates for methods and classes from their FGJ types.
% The templates contain region variables ($\rho$) to denote unknown
% region annotations, and predicate variables ($\varphi$) to denote
% unknown constraints over such region annotations. Free region
% variables are generalized in types (hence, polymorphic). Second, we
% elaborate expressions by introducing region variables to denote
% unknown region arguments in \C{new} expressions, method calls and
% function applications. While elaborating expressions, we make use of
% the computed region type templates to build a system of constraints
% that capture well-formedness requirements and subtyping relationships
% between type templates that must hold (as per the static semantics in
% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
% lift expression elaboration and constraint generation to methods and
% classes. Finally, we solve the constraints to compute solutions to
% region ($\rho$) and predicate ($\varphi$) variables.
%
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
region-annotated definition of \C{Pair} class from
\S\ref{sec:fb-syntax} a region type template for a method with FGJ
type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
accept one region parameter - the allocation region of their
objects.}\footnote{We abuse arrow notation to also represent types of
methods, but unlike function types, there is no allocation region
annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
region variables, and $\varphi_0$ is a fresh predicate variable
denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
$\rho$ is fresh. For a class, a template is computed based on the 
type templates of its fields, constructors and methods. Additional
region and predicate variables are added as necessary. Following template 
of the \C{Pair} class illustrates this point:
% For a class, a template is computed in two steps. In
% the first step, we templatize the types of all its fields, constructor
% arguments, and arguments and return values of all its methods, along
% with the type of its superclass. In the second step, we generalize all
% the free region variables occuring in the templates computed in the
% first step as region parameters of the class. Finally, we add a fresh
% allocation region parameter ($\rhoalloc$) to the list of parameters,
% and introduce a new predicate variable ($\varphi$) to denote unknown
% constraints on region parameters. For example, consider the standard
% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
% respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$. 
%
% The type template for a recursively defined class is computed in two
% steps. First, all recursive occurances of the class among the types of
% its fields are ignored and the class is templatized as if it is a
% non-recursive class. Next, all the recursive occurances are
% templatized with respect to the class template computed in the first
% step, such that their region annotations are \emph{exactly} same as
% the region parameters of the class. Applying this technique on a
% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
% example, results in a template that uses same region variable to
% denote allocation regions of the current node, and the node that the
% current node refers to. 
%
% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
% data stored in the linked list node.  To templatize the
% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
% occurance in the type of $\C{next}$ field, and templatize the type
% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
% template of $\C{data}$ field, we compute the class's template as
% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
% \end{center} Next, we trivially templatize the type of $\C{next}$
% field in the body of the class as
% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
% resulting class represents a linked list with spine in the region
% $\rhoalloc_0$ and data objects in the region $\rho_0$.
% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The type template for for a recursively defined class is computed in
two steps. First, a template is computed by ignoring the recursive
occurrences. Next, all the recursive occurrences are templatized based
on the template computed in first step.  This approach can be extended
to mutually recursive definitions in a straightforward manner. Using
the techniques outlined above, we compute region type templates for
all classes bound in the class table of the $\absof{\FB}$ program
before we proceed to elaborate expressions.

@@

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types occurring in expressions with fresh region
type templates, and (b). explicitly instantiating region parameters
with fresh region variables in method calls and function applications.
This elaboration is performed with respect to the polymorphic type
templates of classes and methods computed as per
\S\ref{sec:fb-templatization}. The elaboration function traverses
expressions in a syntax-directed manner of a type checker, introducing
fresh region type templates for unknown region types, while generating
constraints over region and predicate variables. The precise nature of
generated constraints is explained in \S\ref{sec:fb-constraintsem},
but in summary, they capture the relationships between the type
templates of various subexpressions and their well-formedness.

% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
% this elaboration for a subset of expressions in $\absof{\FB}$
% %, whose corresponding $\FB$ expressions have been ascribed static
% %semantics in Fig.~\ref{fig:fb-staticsem}. 
% $\elabExpr$ is defined under the same context as the expression typing
% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
% and $\env$ retaining their meaning. The function traverses expressions
% in a syntax-directed manner of a type checker, introducing fresh
% region type templates for unknown region types, while generating
% constraints over region and predicate variables. The precise nature of
% generated constraints is explained in \S\ref{sec:fb-constraintsem},
% but in summary, they capture the relationships between the type
% templates of various subexpressions and their well-formedness. Note
% that $\elabExpr$ returns the region type template of the
% subexpression, which is used to generate constraints for the
% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
% shown) used by $\elabExpr$ implement type well-formedness and subtype
% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.

@@

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints. However, we do not
require region type inference to be a whole-program analysis.
Compositionality of type inference is reclaimed by solving the
constraints in a compositional fashion. Concretely, this means that
our constraint solving algorithm visits and solves every constraint
(or, every set of mutually dependent constraints) only once. It
composes computed solutions to solve other constraints that depend on
the solved constraints. Importantly, the failure to solve a dependent
constraint does not result in backtracking. We now describe our
constraint solving algorithm by demonstrating its performance over the
example constraints shown in the previous section. A formal account of
the algorithm can be found in our technical report~\cite{techrep}.
% To simplify the presentation of our algorithm, we assume that are no
% mutually recursive definitions in the source program. Recursive
% definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occurring on the left side (resp. right side) of the turnstile is said
to occur negatively (resp. positively). A validity constraint
\emph{constrains} the set of predicate variables that occur negatively
in the constraint, while it \emph{uses} the set of predicate variables
that occur positively. A constraint is said to be \emph{recursive} if
it constrains and uses a predicate variable.
% A pair of validity constraints are said to be \emph{mutually
% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
% predicate variables that occur with opposite polarities in both the
% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
% predicate variables are also said to be mutually dependent. We call
% the transitive closure of mutual dependency relation as
% \emph{transitive dependency} relation. Like mutual dependency,
% transitive dependency is also extended to predicate variables.

@@

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) variables in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{\cite{techrep} formally justifies the soundness of this
step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}
The constraint $c_{12}$ is a non-recursive constraint of the form
$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
this case). Such constraints are solved for $\varphi$ by computing a
maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
well-formedness condition on $\varphi$, then it is clearly a solution
for $\varphi$, because there cannot exist a weaker solution for
$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
maximally weak $\phisol$ fails the well-formedness condition, then we
establish~\cite{techrep} that it is impossible to satisfy the
constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
and check if it satisfies the well-formedness condition on $\varphi$.
The former problem can be cast as an instance of graph augmentation
problem, as described below.
\vspace*{-0.05in}
\paragraph{Definitions} Given a constraint formula $\phi$, we define
its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
vertices ($V(\phi)$) are free region variables and identifiers in
$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
in\footnote{We alternate between viewing a constraint formula ($\phi$)
as a set of constraints and a conjunction of constraints in this
description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
conjunction of symmetric outlives constraints for the purpose of graph
encoding. Conversely, given a digraph $G$, we define its constraint
encoding $\Phi(G)$ in a straightforward manner. We say that a graph
$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
exist a path between same pair of vertices in $G_1$.

@@

We have so far ignored function types in the context of type
inference. The region type system for $\FB$ admits function types with
higher-rank region polymorphism, allowing region variables to be
quantified in the types of higher-order arguments. The type system
clearly does not have the principal typing property, as demonstrated
by the \C{app2} method below:
\begin{codejava}
bool app2(A x, A y, (A $\rightarrow$ bool) f) 
    { return (f x && f y); }
\end{codejava}
There are two possibilities for the region type of \C{app2}: one,
where it allows \C{x} and \C{y} to be allocated in different regions,
but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
arguments from any region), and second, where it requires \C{x} and
\C{y} to be allocated in the same region, and allows \C{f} to be
monomorphic. None of the two types are more general than the other,
and it is not clear which one is more preferable. The concrete
manifestation of this dilemma is a constraint of form 
$\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
no additional constraints to help determine $\varphi_1$ or $\varphi_2$
before solving this constraint. Since it is always possible to
strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
does not exist unique solution to such constraints. Nonetheless, it is
possible to bias the constraint-solving algorithm to consistently
prefer one solution over the other.
